---
title: "Gene Expression Network Analysis"
author: Christof Seiler
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This Rmd file contains a minimal network analysis pipeline using the packages ``STRINGdb`` and ``BioNet``.

## Install Packages

Install necessary packages from bioconductor repository. Run this code only once to install packages.

```{r eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("STRINGdb")
biocLite("BioNet")
biocLite("DLBCL")
install.packages(c("stringr","igraph"))
```

Load packages.

```{r warning=FALSE,message=FALSE}
library(STRINGdb)
library(BioNet)
library(DLBCL)
library(stringr)
library(igraph)
```

## Load Data

Load example dataset. Replace this with your own list of pvalues.

```{r}
diff_exp = read.csv("Cuffdiff LK_2_5_11_13.csv")
excludeGenesFraction = sum(diff_exp$status == "NOTEST")/length(diff_exp$status)
excludeGenesFraction
diff_exp = diff_exp[!diff_exp$status == "NOTEST",]
dim(diff_exp)
```

## Download Gene Network

Download proteins for human species (code is 9606).

Consider interaciton that are 0.9 confidence. From the STRING website: 

In STRING, each protein-protein interaction is annotated with one or more 'scores'. Importantly, these scores do not indicate the strength or the specificity of the interaction. Instead, they are indicators of confidence, i.e. how likely STRING judges an interaction to be true, given the available evidence. All scores rank from 0 to 1, with 1 being the highest possible confidence. A score of 0.5 would indicate that roughly every second interaction might be erroneous (i.e., a false positive).

```{r}
string_db = STRINGdb$new(version="10", species=9606, score_threshold=900, input_directory="")
```

Check how many proteins are in the database.

```{r}
string_proteins = string_db$get_proteins()
dim(string_proteins)
```

Map gene names to identifiers used in the database.

```{r}
mapped = string_db$map(diff_exp, "gene", removeUnmappedRows = FALSE)
interactions = string_db$get_interactions(mapped$STRING_id)
interactions = data.frame(from=interactions$from,to=interactions$to,combined_score=interactions$combined_score)
dim(interactions)
head(interactions)
```

## Find Subgraph

Fit a Beta-Uniform model.

```{r fig.width=12}
pval = mapped$p_value
names(pval) = mapped$gene
fb = fitBumModel(pval)
fb
```

Set the ``fdr`` parameter which can be interpreted as the FDR of the subgraph. Smaller values will produce a smaller maximum subgraph. You should try a few values (e.g. 0.05, 0.01, 0.001, ...) to obtain a reasonable small subgraph that permits biological interpretation.

```{r}
network = graph_from_data_frame(interactions)
revMappedNames = sapply(names(V(network)),
                        function(string_id)
                          paste(mapped[which(string_id == mapped$STRING_id),]$gene,collapse = "/")
                        )
V(network)$name = unlist(revMappedNames)
scores = scoreNodes(network, fb, fdr=0.0036)
```

Search optimal subgraph.

```{r}
module = runFastHeinz(network, scores)
module
```

Differential expression between ABC and GCB B-cell lymphoma is coloured in red and green, where green shows an upregulation in ACB and red an upregulation in GBC. The shape of the nodes depicts the score: rectangles indicate a negative score, circles a positive score.

```{r fig.width=12, fig.height=12}
par(mfrow=c(1,1))
# logFC = mapped$log2.fold_change.
# names(logFC) = mapped$gene
# plotModule(module, scores = scores, diff.expr = logFC)
plotModule(module, scores = scores)
```
